---
title: "MaizeAndSorghumTranscriptAnalysis"
author: "Nate Korth"
date: "2025-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/GERMsHeatStress/")
library(ggplot2)
library(readxl)
library(DESeq2)
library(apeglm)
library(ggrepel)
library(dplyr)
library(vegan)
library(tidyr)
library(rtracklayer)
library(ggpubr)
library(gplots)
library(igraph)
library(ggraph)
library(ggvenn)
library(UpSetR)
library(stringr)
library(tidyr)
library(edgeR)
library(purrr)
```

#Read in both maize and sorghum genes, this analysis will only consider genes that are orthologous

```{r Read-in files and combine datasets based on orthologous genes}
#Read in MetaData:
MetaData<-read_excel("HotGERMs_SampleSelection.xlsx",sheet = "Sheet2")
#subset only sequenced samples
MetaData<-subset(MetaData,Sequenced=="Yes")
MetaData<-subset(MetaData,Genotype!="BS")
MetaData$Species <- ifelse(MetaData$Genotype == "SAP", "Sorghum", "Maize")

rownames(MetaData)<-MetaData$ID2
MetaData$Genotype <- factor(MetaData$Genotype)
MetaData$Temp2 <- factor(MetaData$Temp2)
MetaData$Soil <- factor(MetaData$Soil)
MetaData$Species<- factor(MetaData$Species)

maize_file_list <- list.files(path = "./FeatureCounts/Annotation3/MaizeFeatureCounts/",pattern=".txt")
maize_path<-"./FeatureCounts/Annotation3/MaizeFeatureCounts"
maize_file_list1<-file.path(maize_path, maize_file_list)
  
# Initialize an empty list to store dataframes
maize_data_list <- list()

# Loop through each file to add feature counts to one file
for (file in maize_file_list1) {
  temp_data <- read.table(file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  base_name <- tools::file_path_sans_ext(basename(file))
  shortened_name <- paste(strsplit(base_name, "_")[[1]][1], collapse = "_")
  colnames(temp_data)[7] <- shortened_name
  maize_data_list[[file]] <- temp_data
}

# Merge all dataframes by the first 6 columns
maize_combined_data <- Reduce(function(x, y) merge(x, y, by = colnames(x)[1:6], all = TRUE), maize_data_list)

sorghum_file_list <- list.files(path = "./FeatureCounts/Annotation3/SorghumFeatureCounts/",pattern=".txt")
sorghum_path<-"./FeatureCounts/Annotation3/SorghumFeatureCounts"
sorghum_file_list1<-file.path(sorghum_path, sorghum_file_list)

# Initialize an empty list to store dataframes
sorghum_data_list <- list()

# Loop through each file to add feature counts to one file
for (file in sorghum_file_list1) {
  temp_data <- read.table(file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  base_name <- tools::file_path_sans_ext(basename(file))
  shortened_name <- paste(strsplit(base_name, "_")[[1]][1], collapse = "_")
  colnames(temp_data)[7] <- shortened_name
  sorghum_data_list[[file]] <- temp_data
}

# Merge all dataframes by the first 6 columns
sorghum_combined_data <- Reduce(function(x, y) merge(x, y, by = colnames(x)[1:6], all = TRUE), sorghum_data_list)

maizegenes<-maize_combined_data$Geneid
sorghumgenes<-gsub("(\\.[^\\.]+){1,2}$","", sorghum_combined_data$Geneid)

# ortholog list made using orthofinder
orthologs<-read.table("./Zm-B73-REFERENCE-NAM-5.0_Zm00001eb.1.protein.tsv",sep="\t",header=T)

# Remove isoform from  gene names in orthologs
orthologs$Maize <- gsub("_P\\d+", "", orthologs$Zm.B73.REFERENCE.NAM.5.0_Zm00001eb.1.protein)

# Split multiple entries into long format
ortho_long <- orthologs %>%
  separate_rows(Maize, sep = ",\\s*") %>%
  separate_rows(Orthologs, sep = ",\\s*") %>%
  dplyr::select(Orthogroup, Maize, Orthologs)

#remove sorghum isoforms
ortho_long$Sorghum <- gsub("(\\.[^\\.]+){2,3}$","", ortho_long$Orthologs)
ortho_long$Orthologs<-NULL

# Filter for genes in both datasets
matched_orthos <- ortho_long %>%
  filter(Maize %in% maizegenes, Sorghum %in% sorghumgenes)


matched_orthos_unique <- matched_orthos %>%
  distinct()

# Preview the result
head(matched_orthos)

# Now filter both maize and sorghum based on genes in the matched ortholog file
maize_filtered <- maize_combined_data %>%
  filter(Geneid %in% matched_orthos_unique$Maize)

sorghum_filtered <- sorghum_combined_data %>%
  mutate(Geneid = gsub("(\\.[^\\.]+){1,2}$", "", Geneid)) %>%
  filter(Geneid %in% matched_orthos_unique$Sorghum)
# Add Orthogroup info to both
maize_annotated <- left_join(maize_filtered, matched_orthos_unique, by = c("Geneid" = "Maize"))
sorghum_annotated <- left_join(sorghum_filtered, matched_orthos_unique, by = c("Geneid" = "Sorghum"))

# Rename Geneid columns for clarity before combining
maize_annotated <- dplyr::rename(maize_annotated, Maize_Geneid = Geneid)

sorghum_annotated <-dplyr::rename(sorghum_annotated,Sorghum_Geneid = Geneid)

# Merge on Orthogroup
combined_data <- full_join(maize_annotated, sorghum_annotated, by = "Orthogroup", suffix = c(".maize", ".sorghum"))

# Preview the result
head(combined_data)

count_cols <- grep("^H\\d+$", names(combined_data), value = TRUE)
count_matrix <- combined_data[, count_cols]
count_matrix$ID<-combined_data$Maize_Geneid
#Remove duplicated rows:
count_matrix <- count_matrix[!duplicated(count_matrix), ]

keep <- rowSums(count_matrix >= 75) >= 4
count_matrix <- count_matrix[keep,]

#reorder the count data to match the metadata
count_matrix1 <-cbind(GeneID=count_matrix$ID,count_matrix[, MetaData$ID2])

#Calculate the average read count of orthologous genes:
mean(colSums(count_matrix1[-c(1)]))
sd(colSums(count_matrix1[-c(1)]))
#write.csv(count_matrix1,"MaizeSorghumOrthologs.csv")
```

```{r PERMANOVA to test terms}
count_cols <- grep("^H\\d+$", names(combined_data), value = TRUE)
count_matrix <- combined_data[, count_cols]
#Remove duplicated rows:
count_matrix <- count_matrix[!duplicated(count_matrix), ]

keep <- rowSums(count_matrix >= 75) >= 4
count_matrix <- count_matrix[keep,]

#reorder the count data to match the metadata
count_matrix <- count_matrix[, MetaData$ID2]

log_counts <- t(log1p(count_matrix))

identical(rownames(log_counts), MetaData$ID2)

# Run PERMANOVA using adonis2 (formula similar to DESeq2 design)
permanova_res <- adonis2(log_counts ~ Genotype * Temp2 * Soil,by="terms", data = MetaData, method = "euclidean")

# View result
print(permanova_res)

```
adonis2(formula = log_counts ~ Genotype * Temp2 * Soil, data = MetaData, method = "euclidean", by = "terms")
                    Df SumOfSqs      R2       F Pr(>F)    
Genotype             2   378744 0.57655 34.7955  0.001 ***
Temp2                1    15603 0.02375  2.8670  0.038 *  
Soil                 1     8000 0.01218  1.4698  0.223    
Genotype:Temp2       2    15304 0.02330  1.4060  0.199    
Genotype:Soil        2    51422 0.07828  4.7242  0.001 ***
Temp2:Soil           1     9748 0.01484  1.7911  0.130    
Genotype:Temp2:Soil  2     9382 0.01428  0.8619  0.501    
Residual            31   168715 0.25683                   
Total               42   656916 1.00000  

PERMANOVA indicates Gentoype:Soil interaction is important for global transcriptional profile

```{r RunDEseq}
#Run a DEseq2 analysis on merged dataset

# Extract only count columns (starting with 'H') — these are sample IDs
count_cols <- grep("^H\\d+$", names(combined_data), value = TRUE)
count_matrix <- combined_data[, count_cols]

combined_data$U_Orthogroup <- make.unique(as.character(combined_data$Orthogroup))

rownames(count_matrix) <- combined_data$U_Orthogroup
#Remove duplicated rows:
count_matrix <- count_matrix[!duplicated(count_matrix), ]

#Final Matrix for Deseq
count_matrix <- count_matrix[rowSums(count_matrix) > 10, ]

#reorder the count data to match the metadata
count_matrix <- count_matrix[, MetaData$ID2]

#Check to make sure meta data and count data are in the same order:
all(rownames(MetaData) == colnames(count_matrix))

#Make DEseq data set:
dds <- DESeqDataSetFromMatrix(countData = count_matrix,
       colData = MetaData,
       design = ~Genotype + Temp2 + Soil + Soil:Temp2)
dds
#Pre filtering Keep rows (genes) that have at least 75 reads in 3 samples 
smallestGroupSize <- 3
keep <- rowSums(counts(dds) >= 80) >= smallestGroupSize
dds <- dds[keep,]

#To test model terms:

#Start wit a full model, then drop interaction terms and us a Likelihood Ratio test to see if dropping the term reduces model fit:
#ddsFULL<- DESeq(dds, test = "LRT", reduced = ~ Genotype + Temp2 + Soil)
#res <- results(ddsFULL)
#head(res,n=1)
#table(res$padj < 0.1)
#InteractionGenes<-subset(as.data.frame(res),padj<=0.1)
#write.csv(InteractionGenes,"InteractionGenesbyLRT.csv")

#Remember to change the model back to how you want it:
ddsA <- DESeq(dds)

normalized_counts <- as.data.frame((counts(ddsA, normalized = TRUE)))

ortholog_map <- combined_data %>%
  dplyr::select(Maize_Geneid, U_Orthogroup) %>%
  filter(!is.na(U_Orthogroup))

#Export a table of orthologs, listed as maize genes and normalized for co-expression analysis: 
normalized_counts1<-normalized_counts %>% 
  tibble::rownames_to_column(var = "U_Orthogroup") %>% 
  left_join(ortholog_map, by = "U_Orthogroup")
normalized_counts1$U_Orthogroup<-NULL
#For orthogroups that contain the same maize gene, sum their expression 
normalized_counts_summed <- as.data.frame(normalized_counts1 %>%
  group_by(Maize_Geneid) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>%
  ungroup())

rownames(normalized_counts_summed)<-normalized_counts_summed$Maize_Geneid
normalized_counts_summed$Maize_Geneid <- NULL
#write.csv(normalized_counts_summed,"MaizeSorghumOrthologsNormalized.csv")

resA <- results(ddsA,contrast = c("Temp2","28C","38C"))
resG1 <- results(ddsA,contrast = c("Genotype","CML103","CML52"))
resG2 <- results(ddsA,contrast = c("Genotype","CML103","SAP"))
resG3 <- results(ddsA,contrast = c("Genotype","SAP","CML52"))
#resG4 <- results(ddsA,contrast = c("Species","Maize","Sorghum"))
resS <- results(ddsA,contrast = c("Soil","Autoclaved","Field"))
#resI<-results(ddsA, name = "GenotypeCML52.Temp238C.SoilField")

#More info or results:
#mcols(resI)$description

#To look at names of variables:
resultsNames(ddsA)

#LogFold Shrinkage
resLFC_A <- lfcShrink(ddsA, coef="Temp2_38C_vs_28C",type = "apeglm")
resLFC_G1 <- lfcShrink(ddsA, contrast = c("Genotype", "CML52", "CML103"), type = "ashr")
resLFC_G2 <- lfcShrink(ddsA, contrast = c("Genotype", "SAP", "CML52"), type = "ashr")
resLFC_G3 <- lfcShrink(ddsA, contrast = c("Genotype", "SAP", "CML103"), type = "ashr")
#resLFC_G4 <- lfcShrink(ddsA, coef="Species_Sorghum_vs_Maize",type = "apeglm")
resLFC_S <- lfcShrink(ddsA, coef="Soil_Field_vs_Autoclaved",type = "apeglm")
#resLFC_I <- lfcShrink(ddsA, coef="GenotypeCML52.Temp238C.SoilField",type = "apeglm")

#Look at number of genes up/down regulated
summary(resA)
summary(resG1)
summary(resG2)
summary(resG3)
#summary(resG4)
summary(resS)
#summary(resI)


```
Summary of LRT results

| Reduced Model Formula                                                     | Genes DE (padj < 0.05) |
| ------------------------------------------------------------------------- | ---------------------- |
| `~ Genotype + Temp2 + Soil + Genotype:Soil + Genotype:Temp2 + Soil:Temp2` | **77**                 |
| `~ Genotype + Temp2 + Soil + Genotype:Soil + Genotype:Temp2`              | **403**                |
| `~ Genotype + Temp2 + Soil + Genotype:Soil + Soil:Temp2`                  | **1,357**              |
| `~ Genotype + Temp2 + Soil + Genotype:Temp2 + Soil:Temp2`                 | **1,055**              |
| `~ Genotype + Temp2 + Soil + Genotype:Soil`                               | **2,019**              |
| `~ Genotype + Temp2 + Soil + Genotype:Temp2`                              | **1,783**              |
| `~ Genotype + Temp2 + Soil + Soil:Temp2`                                  | **2,581**              |
| `~ Genotype + Temp2 + Soil`                                               | **3,236**              |

run table(MetaData$Genotype, MetaData$Temp2, MetaData$Soil) to see:
There's only 2 reps of sorghum at 28C and Field Soil, so any Genotype interaction is underpowered (Probably why there's a large number of DE genes when including those interactions)

because of this underpowing (Result of missing samples) probably best to only include soil x Temp interaction

```{r plot DEseq Results}
#Add a list of genes ID'd in literature to be involved in heatstress:
MaizeLitTable<-read_excel("LiteratureData/DataSummary.xlsx",sheet="AllGenesMaizeV5")
#MaizeLitTable<-subset(MaizeLitTable,Dataset!="Ali2025")
MaizeLitSum <- MaizeLitTable %>%
  group_by(GeneV5) %>%
  summarise(Dataset_Count = n_distinct(Dataset)) %>%
  arrange(desc(Dataset_Count))

ortholog_map <- combined_data %>%
  dplyr::select(Maize_Geneid, U_Orthogroup) %>%
  filter(!is.na(U_Orthogroup))

# Inner join keeps only matching genes and adds Ortholog_U
MaizeLitSum_filtered <- MaizeLitSum %>%
  inner_join(ortholog_map, by = c("GeneV5" = "Maize_Geneid"))

#LFC<-plotMA(resLFC_A,returnData=TRUE)
LFC <- as.data.frame(resLFC_A)

LFC$logBaseMean <- log10(LFC$baseMean)
LFC$Gene<-rownames(LFC)
LFC$pval<-(-1)*log10(LFC$padj)

Lim<-max(abs(LFC$log2FoldChange))


LFC <- LFC %>%
  tibble::rownames_to_column("Orthogroup")

# Join with gene_summary_filtered based on Orthogroup / Ortholog_U
LFC <- LFC %>%
  left_join(MaizeLitSum_filtered %>% dplyr::select(U_Orthogroup, Dataset_Count),
            by = c("Orthogroup" = "U_Orthogroup"))

LFC <- LFC %>%
  left_join(ortholog_map %>% dplyr::select(U_Orthogroup,Maize_Geneid),
            by = c("Orthogroup" = "U_Orthogroup"))

# (Optional) If you want to put the Orthogroup back as rownames
LFC <- LFC %>%
  tibble::column_to_rownames("Orthogroup")
LFC$Dataset_Count[is.na(LFC$Dataset_Count)] <- 0
#Order LFC so high abundance genes are on top:
#LFC <- LFC[order(LFC$Dataset_Count), ]
LFC$Dataset_Count <- as.factor(LFC$Dataset_Count)

#write.csv(LFC,"GenesDEinCML103VsSAP.csv")

#VolcanoPlot (Temp):
ggplot(LFC,aes(x=log2FoldChange,y=pval,color=Dataset_Count))+theme_classic()+xlim((Lim*-1-1),(1+Lim))+
geom_point(alpha=.65,size=1.5)+
  scale_color_manual(values=c("#FAEBDDFF","#F6AE86FF", "#F4815AFF", "#ED4E3EFF","#D3214BFF"))+ #theme(legend.position = "none")+
  #scale_color_gradient(low = "lightgrey", high = "darkred", name = "log10(transcript abundance)")+
  labs(y="-log10 p-value",x = "Log2 Fold Change")+
  annotate("text", x = min(LFC$log2FoldChange, na.rm = TRUE), y = max(LFC$pval, na.rm = TRUE), label = "28C", hjust = 0, size = 6) +
  annotate("text", x = -1*(min(LFC$log2FoldChange, na.rm = TRUE)), y = max(LFC$pval, na.rm = TRUE), label = "38C", hjust = 0, size = 6)+geom_hline(yintercept = 1.577,linetype=2,size=1.1,color="grey33")+
  theme(text=element_text(size=22,family = "sans"))

#ggsave("../Figures/DEseqMaizeSorghumTemp_VolcanoPlot.pdf",width=6,height=6)

process_LFC_results <- function(res_object, padj_threshold = 0.1, lfc_threshold = .1) {
  df <- as.data.frame(res_object)
  df$isDE <- ifelse(!is.na(df$padj) &
                    df$padj < padj_threshold &
                    abs(df$log2FoldChange) > lfc_threshold, TRUE, FALSE)
  df$logBaseMean <- log10(df$baseMean)
  df$Gene <- rownames(df)
  df$pval <- -log10(df$pvalue)
  df$direction <- ifelse(df$isDE & df$log2FoldChange > 0, "Up",
                         ifelse(df$isDE & df$log2FoldChange < 0, "Down", NA))
  df_sig <- subset(df, isDE == TRUE)
  return(df)
}

LFCS<-process_LFC_results(resLFC_S)
LFCT<-process_LFC_results(resLFC_A)
LFCG1<-process_LFC_results(resLFC_G1)
LFCG2<-process_LFC_results(resLFC_G2)
LFCG3<-process_LFC_results(resLFC_G3)

venn_data2.1 <- list(
  Soil = LFCS$Gene,
  Temperature = LFCT$Gene,
  CML52vsCML103 = LFCG1$Gene,
  SAPvsCML52 = LFCG2$Gene,
  SAPvsCML103 = LFCG3$Gene
)
#venn_data <- euler(venn_data2.1)

#pdf("./Figures/MaizeTranscriptVenn1.pdf", width = 6, height = 6)
#plot(venn_data,
#     fills = c("brown4", "orange3", "green4","green3","green1"),
#     labels = list(font = 4),
#     quantities = TRUE)

#dev.off()

# Make an upset plot:
# Create a combined unique gene list
soil_de_genes <- LFCS$Gene[LFCS$isDE]
temp_de_genes <- LFCT$Gene[LFCT$isDE]
g1_de_genes   <- LFCG1$Gene[LFCG1$isDE]
g2_de_genes   <- LFCG2$Gene[LFCG2$isDE]
g3_de_genes   <- LFCG3$Gene[LFCG3$isDE]
de_genotype <- unique(c(g1_de_genes, g2_de_genes, g3_de_genes))
de_all3 <- Reduce(intersect, list(temp_de_genes, soil_de_genes, de_genotype))


# Create a combined unique list of ONLY significant DEGs
all_de_genes <- unique(c(soil_de_genes, temp_de_genes, g1_de_genes, g2_de_genes, g3_de_genes))

# Build the binary matrix based on these significant genes
upset_data <- data.frame(
  Gene = all_de_genes,
  Soil = as.integer(all_de_genes %in% soil_de_genes),
  Temperature = as.integer(all_de_genes %in% temp_de_genes),
  CML52vsCML103 = as.integer(all_de_genes %in% g1_de_genes),
  SAPvsCML52 = as.integer(all_de_genes %in% g2_de_genes),
  SAPvsCML103 = as.integer(all_de_genes %in% g3_de_genes)
)

# Remove "Gene" column and plot
upset_matrix <- upset_data
upset_matrix$Gene<-NULL


upset(upset_matrix, sets = c("Soil", "Temperature", "CML52vsCML103","SAPvsCML52","SAPvsCML103"),
  sets.bar.color = "firebrick", order.by = "freq",keep.order = TRUE,text.scale = c(2.2, 2.2, 1.9, 1.8, 1.8, 1.7))

#Plot a stacked bar to add to upset plot:
LFCS$comparison <- "Soil"
LFCT$comparison <- "Temperature"
LFCG1$comparison <- "CML52 vs CML103"
LFCG2$comparison <- "SAP vs CML52"
LFCG3$comparison <- "SAP vs CML103"
all_DEGs <- bind_rows(LFCS, LFCT, LFCG1, LFCG2, LFCG3)

deg_summary <- all_DEGs %>%
  group_by(comparison, direction) %>%
  summarise(n = n(), .groups = "drop")

ggplot(deg_summary, aes(x = comparison, y = n, fill = direction)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Up" = "firebrick", "Down" = "steelblue3")) +
  labs(x = "Comparison",y = "Number of DEGs",
    fill = "Direction") + theme_void() +theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top")
#ggsave("./Figures/UpDownGenesStackedBar.pdf",width = 4,height = 7)

#Calculate the number of DE genes supported by literature: 
LFCA2<-process_LFC_results(LFC)
LFCA2$Dataset_Count<-as.numeric(LFCA2$Dataset_Count)
LFCA2_Supported<-subset(LFCA2,Dataset_Count>=4)
LFCA2_SupportedT<-subset(LFCA2_Supported,isDE==TRUE)
LFCA2T<-subset(LFCA2,isDE==TRUE)
```
Plot upsetPlot:
```{r MaizeTranscriptUpset1, include=FALSE}

pdf("Figures/MaizeTranscriptUpset1.pdf", width = 14, height = 8)

upset(upset_matrix, sets = c("Soil", "Temperature", "CML52vsCML103","SAPvsCML52","SAPvsCML103"),
  sets.bar.color = "firebrick", order.by = "freq",keep.order = TRUE,text.scale = c(2.2, 2.2, 1.9, 1.8, 1.8, 1.7))

dev.off()

```


```{r GO tests}
library(biomaRt)

#Pull all annotations:
# Set up Ensembl Plants mart for maize
ensembl <- useMart("plants_mart", dataset = "zmays_eg_gene", host = "https://plants.ensembl.org")

# List of all maize genes that have orthologs
MaizeOrthos1<-unique(rownames(dds))
MaizeGenes1<-combined_data %>%
  filter(U_Orthogroup %in% MaizeOrthos1) %>%
  pull(Maize_Geneid) %>%
  unique() %>%
  na.omit()

all_annotations <- getBM(
  attributes = c("ensembl_gene_id","external_gene_name","go_id","namespace_1003","name_1006"),
  filters = "ensembl_gene_id",values = MaizeGenes1,mart = ensembl)

# Filter to only biological_process GO terms
go_bg <- all_annotations %>%
  filter(!is.na(name_1006), namespace_1003 == "biological_process")


get_go_enrichment_from_orthogroups <- function(sig_genes, combined_data, mart, comparison_name, go_bg) {
  # 1. Map orthogroups to maize gene IDs
  gene_ids <- merge(
    data.frame(Orthogroup = sig_genes),
    combined_data,
    by.x = "Orthogroup",
    by.y = "U_Orthogroup",
    all.x = TRUE
  )$Maize_Geneid %>% unique() %>% na.omit()
  
  # 2. Retrieve raw GO annotations for DE genes
  go_df <- getBM(
    attributes = c("chromosome_name", "transcript_length", "ensembl_gene_id",           "external_gene_name", "go_id", "namespace_1003", "name_1006"),
    filters = "ensembl_gene_id",values = gene_ids,mart = mart)
  
  
  
  # 4. Background GO counts
  bg_counts <- go_bg %>%
    filter(!is.na(name_1006), namespace_1003 == "biological_process") %>%
    distinct(ensembl_gene_id, name_1006) %>%
    count(name_1006, name = "bg_count") %>%
  filter(bg_count > 5)

  # 5. DE GO counts
  de_counts <- go_df %>%
    filter(!is.na(name_1006), namespace_1003 == "biological_process") %>%
    distinct(ensembl_gene_id, name_1006) %>%
    count(name_1006, name = "de_count") %>%
  filter(de_count > 5)
  # 3. Total background gene count
  N <- length(unique(bg_counts$name_1006))
  
  # 6. Enrichment test (Fisher's exact)
  enrichment_df <- full_join(de_counts, bg_counts, by = "name_1006") %>%
    replace_na(list(de_count = 0, bg_count = 0)) %>%
    rowwise() %>%
    mutate(
      total_de = sum(de_count),
      total_bg = N,
      pval = phyper(de_count - 1, bg_count, total_bg - bg_count, total_de, lower.tail = FALSE)
    ) %>%
    ungroup() %>%
    mutate(
      adj_pval = p.adjust(pval, method = "BH"),
      Comparison = comparison_name
    ) %>%
    arrange(adj_pval)

  return(list(enrichment = enrichment_df, annotations = go_df))
}

result_temp <- get_go_enrichment_from_orthogroups(de_temp, combined_data, ensembl, "Temperature", go_bg)
enriched_go <- result_temp$enrichment

```

#Same thing with TopGO

```{r topgo}
library(topGO)
library(tibble)
go_bg <- all_annotations %>%
  filter(!is.na(name_1006), namespace_1003 == "biological_process") %>%
  filter(go_id > 10)

get_go_enrichment_from_orthogroups <- function(sig_genes, combined_data, mart, comparison_name, go_bg) {

  # 1. Map orthogroups to maize gene IDs
  gene_ids <- merge(
    data.frame(Orthogroup = sig_genes),
    combined_data,
    by.x = "Orthogroup",
    by.y = "U_Orthogroup",
    all.x = TRUE
  )$Maize_Geneid %>% unique() %>% na.omit()

  # 2. Create gene universe
  all_genes <- unique(go_bg$ensembl_gene_id)
  #gene_list <- factor(as.integer(all_genes %in% gene_ids))
  gene_list <- factor(as.integer(all_genes %in% gene_ids), levels = c("0", "1"))
  names(gene_list) <- all_genes

  # 3. Create gene-to-GO mapping
  geneID2GO <- go_bg %>%
    filter(!is.na(go_id), namespace_1003 == "biological_process") %>%
    group_by(ensembl_gene_id) %>%
    summarise(go_terms = list(unique(go_id))) %>%
    deframe()

  # 4. Build topGO data object
  GOdata <- new("topGOdata",
                ontology = "BP",
                allGenes = gene_list,
                annot = annFUN.gene2GO,
                gene2GO = geneID2GO,
                nodeSize = 5)

  # 5. Run enrichment test
  resultFisher <- runTest(GOdata, algorithm = "weight01", statistic = "fisher")

  # 6. Extract results table
  enrichment_df <- GenTable(
    GOdata,
    Fisher = resultFisher,
    orderBy = "Fisher",
    topNodes = length(score(resultFisher))) %>%
    filter(Annotated >= 2) %>%
    rename(name_1006 = Term,go_id = GO.ID,pval = Fisher) %>%
    mutate(pval = as.numeric(pval),
      Significant = as.numeric(Significant),
      Expected = as.numeric(Expected),
      # Calculate Log2 Fold Enrichment
      Log2FoldEnrichment = log2(Significant / Expected),
      adj_pval = p.adjust(pval, method = "fdr"),
      Comparison = comparison_name) %>%
    arrange(adj_pval)

  # 7. Return both enrichment and raw annotations
  return(list(enrichment = enrichment_df, annotations = geneID2GO))
}


#result_temp <- get_go_enrichment_from_orthogroups(de_temp, combined_data, ensembl, "Temperature", go_bg)
#T_enriched_go <- result_temp$enrichment

go_results <- list(
  Intersect = get_go_enrichment_from_orthogroups(de_all3, combined_data, ensembl, "Intersect", go_bg),
  Temperature   = get_go_enrichment_from_orthogroups(temp_de_genes, combined_data, ensembl, "Temperature", go_bg),
  Soil= get_go_enrichment_from_orthogroups(soil_de_genes, combined_data, ensembl, "Soil", go_bg),
  CML52vsCML103 = get_go_enrichment_from_orthogroups(g1_de_genes,   combined_data, ensembl, "CML52vsCML103", go_bg),
  SAPvsCML52     = get_go_enrichment_from_orthogroups(g2_de_genes,   combined_data, ensembl, "SAPvsCML52", go_bg), 
  SAPvsCML103   = get_go_enrichment_from_orthogroups(g3_de_genes,   combined_data, ensembl, "SAPvsCML103", go_bg))


# Extract and combine significant terms
significant_go_df <- map_dfr(
  go_results,
  ~ .x$enrichment %>% 
    filter(pval < .05),
  .id = "Comparison")
#write.csv(significant_go_df,"SignificantGOTermsPlant.csv")
```

```{r plot topgo results}
# First for just the intersect:
IntersectGO<-subset(significant_go_df,Comparison=="Intersect")
IntersectGO<-subset(IntersectGO,Significant>=4)

# Create long format for plotting both Observed and Expected
go_long <- IntersectGO %>%
  dplyr::select(name_1006, Expected, Significant) %>%
  pivot_longer(cols = c(Expected, Significant),
               names_to = "Type",
               values_to = "Count")

# Segment data (wide form again) to connect Expected -> Observed
segments_df <- IntersectGO %>%
  dplyr::select(name_1006, Expected, Significant)

# Plot
ggplot(go_long, aes(x = reorder(name_1006, Count), y = Count, color = Type)) +
    coord_flip() + theme_classic(base_size = 14) +
  labs(x = "GO Term",y = "Gene Count",color = "Count Type") +
  scale_color_manual(values = c("Expected" = "gray40", "Significant" = "firebrick")) +
  geom_segment(data = segments_df,
               aes(x = name_1006, xend = name_1006, y = Expected, yend = Significant),
               inherit.aes = FALSE,
               color = "gray20", linewidth = 0.8)+geom_point(size = 4) +
  theme(axis.text.y = element_text(size = 13))+
  theme(text=element_text(size=20,family = "sans"))

#ggsave("./Figures/IntersectGos.pdf",width=10,height=5)

```
Same GO enrichment plot but show log fold enrichment
```{r}
# Filter for the Intersect terms with enough signal
PlotData <- significant_go_df %>%
  filter(Comparison == "Intersect", Significant >= 5) %>%
  # Optionally filter for only the top 15-20 most enriched terms for clarity
  slice_min(order_by = pval, n = 20)

ggplot(PlotData, aes(x = reorder(name_1006, Log2FoldEnrichment), y = Log2FoldEnrichment, fill = Significant)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  scale_fill_gradient(low = "#3C1A42FF", high = "#ED4E3EFF") + # Color by significance
  theme_classic(base_size = 14) +
  labs(
    x = "Biological Process",
    y = "Log2 Fold Enrichment (Observed/Expected)",
    fill = "Gene Number"
  ) +
  theme(
    text = element_text(family = "sans", size = 18),
    axis.text.y = element_text(size = 12)
  )

ggsave("../Figures/PlantIntersectGOTermsLFC.jpeg",dpi=600,width =10,height=5)
```



```{r convert sorghum geneIDs}
SorgGenes0<-read_excel("LiteratureData/DataSummary.xlsx",sheet="Sorghum2Maize")

sorgGenes1<-unique(merge(SorgGenes0,ortho_long,by.x="SorghumGene",by.y="Sorghum"))
write.csv(sorgGenes1,"Sorghum2MaizeGenes.csv")

MaizeGenes0<-read_excel("LiteratureData/DataSummary.xlsx",sheet="Maize2Maize")
MaizeTable<-read.csv("C:/Users/njkorth/Data/Genomes/maize_assembly_gene_model_conversion.csv")

MaizeGenes1<-unique(merge(MaizeGenes0,MaizeTable,by.x="V4",by.y="B73_GRAMENE_4.0"))
write.csv(MaizeGenes1[1:2],"Maize2MaizeGenes.csv")

```

