---
title: "MaizeAndMicrobeCoexpression"
author: "Nate Korth"
date: "2025-08-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
knitr::opts_knit$set(root.dir = "~/GERMsHeatStress/")
library(tidyverse)
library(reshape2)
library(vegan)
library(igraph)
library(ggraph)
library(randomForest)
library(glmnet)
library(caret)
library(compositions)
library(readxl)
library(mixOmics)
library(tidygraph)
library(scales)
```

Load and Prepare Data

```{r}
setwd("~/GERMsHeatStress/")

# Load metadata
MetaData <- read_excel("HotGERMs_SampleSelection.xlsx", sheet = "Sheet2")
rownames(MetaData) <- MetaData$ID2

# Keep only sequenced and exclude Genotype "BS"
MetaData <- MetaData %>%
  filter(Sequenced == "Yes", Genotype != "BS") %>%
  mutate(
    Biomass_Shoot = as.numeric(`Biomass(Shoot)`),
    Species = ifelse(Genotype == "SAP", "Sorghum", "Maize")
  )

# Load Plant data (normalized counts)
Plant <- read.csv("MaizeSorghumOrthologsNormalized.csv", row.names = 1)

PlantPathway<-read.csv("PlantExpression_byKEGGPathway.csv", row.names = 1)
rownames(PlantPathway) <- gsub(" - Zea mays \\(maize\\)", "", rownames(PlantPathway))

# Filter Plant: keep genes with counts >= 70 in more than 3 samples
Plant2 <- Plant %>% dplyr::select(where(is.numeric)) %>% filter(rowSums(. >= 250) > 3)

# Load Bacterial pathways data and filter out mammalian pathways
exclude_paths <- c("Secondary bile acid biosynthesis","Antigen processing and presentation","Apelin signaling pathway","Apoptosis","Retrograde endocannabinoid signaling","Basal transcription factors","Complement and coagulation cascades","Efferocytosis","Fc gamma R-mediated phagocytosis","Ferroptosis","Gastric acid secretion","Glucagon signaling pathway","Neuroactive ligand-receptor interaction","Neutrophil extracellular trap formation","Oxytocin signaling pathway","Parathyroid hormone synthesis","Phagosome","Platinum drug resistance","Toxoplasmosis","Virion - Bacteriophage lambda","Human cytomegalovirus infection","Drug metabolism - cytochrome P450","Primary bile acid biosynthesis")
#Single Genes
bacteria <- read.csv("AnnotationOutputs/GeneCounts/Microbe_Genes_collapsed_counts.csv", row.names = 1)
#Collapsed at Pathway
bacteriaPaths<-read.csv("AnnotationOutputs/GeneCounts/Microbe_Pathways_collapsed_EC_counts.csv")
bacteriaPaths$X<-NULL

########TODO do network with pathways

bacteria<-bacteriaPaths

bacteria <- bacteria[!bacteria$desc %in% exclude_paths, ]

bacteria <- bacteria %>%
  mutate(desc = ifelse(desc == 
           "Biosynthesis of various antibiotics; Including: Kanosamine biosynthesis, Aurachin biosynthesis, Bacilysin biosynthesis, Puromycin biosynthesis, Dapdiamides biosynthesis, Fosfomycin biosynthesis, Cremeomycin biosynthesis, Fumagillin biosynthesis, Pentalenolactone biosynthesis, Terpentecin biosynthesis, Roseoflavin biosynthesis, Cycloserine biosynthesis","Biosynthesis of antibiotics",desc))
bacteria <- bacteria %>%
  mutate(desc = ifelse(desc == 
           "Biosynthesis of various other secondary metabolites; Including: Ditryptophenaline biosynthesis, Fumiquinazoline D biosynthesis, Paerucumarin biosynthesis, Staphyloferrin B biosynthesis, Cyclooctatin biosynthesis, Lovastatin biosynthesis, Grixazone biosynthesis, Staphyloferrin A biosynthesis, Ethynylserine biosynthesis, Aerobactin biosynthesis","Secondary metabolites biosynthesis A",desc))
bacteria <- bacteria %>%
  mutate(desc = ifelse(desc == 
           "Biosynthesis of various plant secondary metabolites; Including: Crocin biosynthesis, Ginsenoside biosynthesis, Saponin adjuvant biosynthesis, Cannabidiol biosynthesis, Mugineic acid biosynthesis, Pentagalloylglucose biosynthesis, Benzoxazinoid biosynthesis, Gramine biosynthesis, Coumarin biosynthesis, Furanocoumarin biosynthesis, Hordatine biosynthesis, Podophyllotoxin biosynthesis","Secondary metabolites biosynthesis B",desc))
bacteria <- bacteria %>%
  mutate(desc = ifelse(desc == 
           "Biosynthesis of various alkaloids; Including: Cucurbitacin biosynthesis, Solanine and tomatine biosynthesis, Ephedrine biosynthesis, Capsaicin biosynthesis, Acridone alkaloid biosynthesis","Alkaloid Biosynthesis",desc))
bacteria <- bacteria %>%
  mutate(desc = ifelse(desc == 
           "Biosynthesis of 12-, 14- and 16-membered macrolides; Including: Tylosin biosynthesis, Mycinamicin biosynthesis, Erythromycin biosynthesis, Oleandomycin biosynthesis, Pikromycin/methymycin biosynthesis, Avermectin biosynthesis","Macrolides Biosynthesis",desc))




rownames(bacteria)<-bacteria$desc
# Filter bacteria rows: keep pathways with counts >=150 in more than 3 samples
bacteria2 <- bacteria %>% dplyr::select(where(is.numeric)) %>% filter(rowSums(. >= 150) > 3)

# CLR transform bacteria data
bacteria2_clr <- as.data.frame(apply(bacteria2 + 1, 2, clr))
colnames(bacteria2_clr)<- gsub("\\.emapper\\.annotations$","",colnames(bacteria2_clr))

# Ensure common samples across datasets
common_cols <- intersect(colnames(Plant2), colnames(bacteria2_clr))

Plant2 <- as.matrix(Plant2[, common_cols])
bacteria2 <- as.matrix(bacteria2_clr[, common_cols])



# Scale for mantel and regression analyses
Plant2_Scaled <- scale(Plant2, center = TRUE, scale = TRUE)
bacteria2_Scaled <- scale(bacteria2, center = TRUE, scale = TRUE)

# Keep unscaled for network and RF
Plant_raw <- Plant2
bacteria_raw <- bacteria2

```
1. Mantel Test and PERMANOVA
This section calculates distance matrices on scaled data, runs Mantel tests and PERMANOVAs to assess relationships between plant transcriptome and microbial pathways.

```{r}
# Calculate Euclidean distances
distPlant <- dist(t(Plant2_Scaled), method = "euclidean")
distBacteria <- dist(t(bacteria2_Scaled), method = "euclidean")

# Mantel test with Spearman correlation, 5000 permutations
mantel_result <- vegan::mantel(distPlant, distBacteria, method = "spearman", permutations = 5000)
print(mantel_result)

# Prepare sample pairs for grouping
sample_names <- attr(distPlant, "Labels")
sample_pairs <- combn(sample_names, 2, simplify = FALSE)

rownames(MetaData) <- MetaData$ID2

soil1 <- vapply(sample_pairs, function(x) as.character(MetaData[x[1], "Soil"]), character(1))
soil2 <- vapply(sample_pairs, function(x) as.character(MetaData[x[2], "Soil"]), character(1))

soil_group <- ifelse(soil1 == soil2, "Same Soil", "Different Soil")

temp1 <- vapply(sample_pairs, function(x) as.character(MetaData[x[1], "Temp2"]), character(1))
temp2 <- vapply(sample_pairs, function(x) as.character(MetaData[x[2], "Temp2"]), character(1))
temp_group <- ifelse(temp1 == temp2, "Same Temperature", "Different Temperature")

species1 <- vapply(sample_pairs, function(x) as.character(MetaData[x[1], "Species"]), character(1))
species2 <- vapply(sample_pairs, function(x) as.character(MetaData[x[2], "Species"]), character(1))
species_group <- ifelse(species1 == species2, "Same Species", "Different Species")

mantel_df <- data.frame(
  PlantDist = as.vector(distPlant),
  BacteriaDist = as.vector(distBacteria),
  SoilGroup = soil_group,
  TempGroup = temp_group,
  SpeciesGroup = species_group
)

# PERMANOVAs
adonis_bact <- adonis2(distBacteria ~ Soil + Species + Temp2 + Soil:Temp2, data = MetaData, permutations = 5000, by = "terms")
adonis_plant <- adonis2(distPlant ~ Soil + Species + Temp2 + Soil:Temp2, data = MetaData, permutations = 5000, by = "terms")

# Format PERMANOVA output for plotting subtitle
format_perm <- function(res) {
  res <- res[!rownames(res) %in% c("Residual", "Total"), ]
  paste0(rownames(res), ": ", round(res$R2, 2), " (p=", signif(res$`Pr(>F)`, 2), ")", collapse = " | ")
}

stats_bact <- format_perm(adonis_bact)
stats_plant <- format_perm(adonis_plant)

# Plot Mantel scatterplot with groups
mantelplot<-ggplot(mantel_df, aes(x = PlantDist, y = BacteriaDist, color = SoilGroup, shape = SpeciesGroup)) +
  geom_point(alpha = 0.65, size = 2.5) +
  scale_color_manual(values = c("Same Soil" = "brown", "Different Soil" = "blue")) +
  theme_classic(base_size = 14) +
  labs(title = paste0("Mantel r = ", round(mantel_result$statistic, 3), " | p = ", round(mantel_result$signif, 4)),subtitle = paste0("PERMANOVA (Bacteria): ", stats_bact, "\nPERMANOVA (Plant): ", stats_plant),x = "Plant Distance",y = "Microbial Distance",color = "Soil Comparison")+theme(axis.text = element_text(size = 18, family = "sans"),axis.title = element_text(size = 18, family = "sans"))

ggsave("./Figures/MantelResults4.pdf",plot=mantelplot, width = 12, height = 7)
print(mantelplot)
```

2. Network Analysis of Strong Gene-Pathway Correlations
Build a network from strong Spearman correlations (>0.7) between raw gene and pathway abundance data.
```{r}
# Transpose raw matrices: samples x features
gene_mat <- t(Plant_raw)
bact_mat <- t(bacteria_raw)

# Calculate Spearman correlations
cor_matrix <- cor(gene_mat, bact_mat, method = "spearman", use = "pairwise.complete.obs")

# Extract strong correlations
cor_df <- reshape2::melt(cor_matrix)
colnames(cor_df) <- c("PlantGene", "MicrobeGene", "Correlation")
edges <- subset(cor_df, abs(Correlation) >= 0.65)

# Build graph object
nodes <- data.frame(name = unique(c(edges$PlantGene, edges$MicrobeGene)))
nodes$type <- ifelse(nodes$name %in% colnames(gene_mat), "PlantGene", "MicrobeGene")

graph <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
graph <- delete.vertices(graph, degree(graph) == 0)

# compute components and their sizes
comp <- components(graph)         # list with $membership and $csize
membership <- comp$membership     # integer vector length = vcount(graph)
csize <- comp$csize               # vector of component sizes indexed by comp id

# choose a minimum component size to keep (tweak as needed)
min_comp_size <- 2

# logical vector: TRUE for vertices in components larger than min_comp_size
keep_vertex_logical <- csize[membership] > min_comp_size

# convert logical into vertex sequence to keep
keep_vertices <- V(graph)[keep_vertex_logical]

# build subgraph containing only desired components
graph_sub <- induced_subgraph(graph, vids = keep_vertices)

# quick check: distribution of component sizes in subgraph
table(components(graph_sub)$csize)

tg <- as_tbl_graph(graph_sub) %>%
  mutate(degree = centrality_degree(), component_id = group_components(),   
         component_size = graph_size()) %>%
  filter(component_size > 2) %>%
  #mutate(label = ifelse(type == "MicrobeGene" & degree > 5, name, NA))
  mutate(label = ifelse(degree > 10, name, NA))
table(components(tg)$csize)

# Plot network with layout "fr"
set.seed(47)
ggraph(tg, layout = "kk") +
  geom_edge_link(aes(edge_alpha = abs(Correlation), edge_width = abs(Correlation), color = Correlation)) +
  geom_node_point(aes(color = type, size = degree)) +
  geom_node_text(aes(label = label), repel = TRUE, size = 6) +
  scale_edge_color_gradient2(low = "#006401", mid = "white", high = "#F5BE61", midpoint = 0) +
  scale_edge_alpha(range = c(0.5, 1)) +
  scale_edge_width(range = c(0.4, 2)) +
  scale_color_manual(values = c("PlantGene" = "cyan", "MicrobeGene" = "purple4")) +
  scale_size_continuous(range = c(2, 8)) +
  theme_void()
#ggsave("./Figures/NetworkGraph_MicrobePathsVPlantGenes.pdf", width = 11, height = 17)

#Same Plot in Black
set.seed(47)
ggraph(tg, layout = "fr") +
  geom_edge_link(aes(edge_alpha = abs(Correlation), edge_width = abs(Correlation), color = Correlation)) +
  geom_node_point(aes(color = type, size = degree)) +
  #geom_node_text(aes(label = label), repel = TRUE, size = 4, color = "white") +
  scale_edge_color_gradient2(low = "#006401", mid = "white", high = "#F5BE61", midpoint = 0) +
  scale_edge_alpha(range = c(0.5, 1)) +
  scale_edge_width(range = c(0.4, 2)) +
  scale_color_manual(values = c("PlantGene" = "cyan", "MicrobeGene" = "purple4")) +
  scale_size_continuous(range = c(2, 8)) +
  theme_void()+theme(
    plot.background = element_rect(fill = "#171717", color = NA),
    panel.background = element_rect(fill = "#171717", color = NA),
    legend.background = element_rect(fill = "#171717", color = NA),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white")
  )
#ggsave("./Figures/NetworkGraph_MGenesVPlantGenes_black.pdf", width = 12, height = 10)
# Extract top connected genes and pathways for further analysis
node_info <- as_tibble(tg) %>%
  arrange(desc(degree))

Gene_Network <- node_info %>% filter(type == "PlantGene") %>% slice_max(degree, n = 30)
Pathways_Network <- node_info %>% filter(type == "MicrobeGene") %>% slice_max(degree, n = 30)
```



3. Elastic Net Regression (Biomass & HeatStress)
Apply elastic net regression using scaled data to predict biomass and heat stress separately.

```{r}
#Add tag for plant or microbe: 
rownames(Plant2_Scaled)  <- paste0("Plant_", rownames(Plant2_Scaled))
rownames(bacteria2_Scaled) <- paste0("Microbe_", rownames(bacteria2_Scaled))
# Prepare combined matrix for regression (scaled data)
X_scaled <- cbind(t(Plant2_Scaled), t(bacteria2_Scaled))

# Align metadata with samples in X
meta_aligned <- MetaData[match(rownames(X_scaled), MetaData$ID2), ]

# Define function to run elastic net and plot results for a given response variable
run_elastic_net <- function(y_var, X, meta, response_label) {
  y_vec <- meta[[y_var]]
  features <- colnames(X)
  
  # Filter features by Spearman correlation p <= 0.05
  pvals <- sapply(features, function(f) cor.test(X[, f], y_vec, method = "spearman")$p.value)
  features_to_keep <- features[pvals <= 0.1]
  
  X_filtered <- as.matrix(X[, features_to_keep, drop = FALSE])
  
  set.seed(42)
  cv_fit <- cv.glmnet(X_filtered, y_vec, alpha = 0.5, standardize = TRUE)
  
  # Extract coefficients at best lambda
  best_lambda <- cv_fit$lambda.min
  coefs <- coef(cv_fit, s = best_lambda)
  coefs_df <- data.frame(
    Feature = rownames(coefs)[which(coefs != 0)],
    Coefficient = coefs[which(coefs != 0)],
    stringsAsFactors = FALSE
  )
  coefs_df <- subset(coefs_df, Feature != "(Intercept)")
  
  # Annotate type
  #coefs_df$Type <- ifelse(coefs_df$Feature %in% rownames(Plant2), "Gene", "Pathway")
  coefs_df$Type <- ifelse(grepl("^Plant_", coefs_df$Feature), "Plant", "Microbe")
  
  coefs_df$Feature <- gsub("^Plant_|^Microbe_", "", coefs_df$Feature)
  
  # Order for plotting
  coefs_df <- coefs_df %>%
    mutate(Type = factor(Type, levels = c("Microbe", "Plant"))) %>%  # Pathway first
    arrange(Type, desc(Coefficient)) %>%
    mutate(Feature = factor(Feature, levels = rev(unique(Feature))))
  coefs_df1<-subset(coefs_df,abs(Coefficient)>=.03)
  
  # Plot
  p <- ggplot(coefs_df1, aes(x = Coefficient, y = Feature, fill = Coefficient > 0.01)) +
    geom_col(show.legend = FALSE) +
    scale_fill_manual(values = c("firebrick", "steelblue")) +
    scale_y_discrete(labels = function(x) sub("^Microbe_|^Plant_", "", x)) +
    labs(x = paste0("Effect on ", response_label), y = NULL, title = paste("Elastic Net Regression -", response_label)) +
    geom_hline(yintercept = sum(coefs_df1$Type == "Gene") + 0.5, color = "black") +
    theme_minimal(base_size = 15) +
    theme(panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank())
  
  ggsave(paste0("./Figures/ElasticNetPATHs_", response_label, ".pdf"), plot = p, width = 13, height = 7)
  print(p)
  return(coefs_df)
}

# Run on Biomass
coefs_biomass <- run_elastic_net("Biomass_Shoot", X_scaled, meta_aligned, "Biomass")

# Run on HeatStress (make sure HeatStress is numeric and present in metadata)
meta_aligned$HeatStress <- as.numeric(meta_aligned$HeatStress)
coefs_heatstress <- run_elastic_net("HeatStress", X_scaled, meta_aligned, "HeatStress")

```
4. Random Forest (Biomass & HeatStress)
Apply Random Forest using raw (unscaled) data to predict biomass and heat stress.
```{r}
# Add prefixes for tracking
rownames(Plant_raw) <- paste0("Plant_", rownames(Plant_raw))
rownames(bacteria_raw) <- paste0("Microbe_", rownames(bacteria_raw))

# Prepare data (raw, filtered features from elastic net step)
X_raw <- cbind(t(Plant_raw), t(bacteria_raw))

# Function to run RF and plot importance
run_rf <- function(y_var, X, meta, response_label) {
  # Align metadata
  meta_aligned <- meta[match(rownames(X), meta$ID2), ]
  y_vec <- as.numeric(meta_aligned[[y_var]])
  
  # Spearman correlation filtering (p <= 0.05)
  pvals <- sapply(colnames(X), function(f) {
    cor.test(X[, f], y_vec, method = "spearman")$p.value
  })
  significant_features <- colnames(X)[pvals <= 0.1]
  
  X_filtered <- as.data.frame(X[, significant_features, drop = FALSE])
  
  set.seed(42)
  train_control <- trainControl(method = "cv", number = 10)
  
  rf_cv <- train(
    x = X_filtered,
    y = y_vec,
    method = "rf",
    trControl = train_control,
    importance = TRUE,
    ntree = 2000
  )
  
  importance_df <- varImp(rf_cv)$importance
  importance_df$Feature <- rownames(importance_df)
  #importance_df$Type <- ifelse(importance_df$Feature %in% rownames(Plant_raw), "Gene", "Pathway")
   # Annotate Type based on prefix
  importance_df$Type <- ifelse(grepl("^Plant_", importance_df$Feature), "Plant", "Microbe")
  # Clean feature names
  importance_df$Feature <- gsub("^Plant_|^Microbe_", "", importance_df$Feature)
  
  # Select top features for plotting
  top_features <- importance_df %>%
  arrange(desc(Overall)) %>%
  group_by(Type) %>%
  slice_max(Overall, n = 15) %>%
  ungroup() %>%
  mutate(PlotFeature = paste(Type, Feature, sep = "_"),   # unique for factor
         PlotFeature = factor(PlotFeature, levels = rev(unique(PlotFeature))),
         Label = gsub("^Plant_|^Microbe_", "", Feature)) 
  
  p <- ggplot(top_features, aes(x = PlotFeature, y = Overall, fill = Type)) +
    geom_col(show.legend = FALSE) +
    scale_fill_manual(values = c("Microbe" = "orange", "Plant" = "green4")) +
    labs(title = paste("Random Forest Variable Importance (10-fold CV) -", response_label),
         x = NULL, y = "Variable Importance") +
    theme_classic() +
    theme(text = element_text(size = 16, family = "sans"),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_continuous(expand = c(0, 0))+
  scale_x_discrete(labels = top_features$Label) 
  
  ggsave(paste0("./Figures/RandomForest_Paths", response_label, ".pdf"), plot = p, width = 12, height = 7)
  
  return(importance_df)
}

# Run RF for Biomass
rf_biomass <- run_rf("Biomass_Shoot", X_raw, MetaData, "Biomass")

# Run RF for HeatStress
rf_heatstress <- run_rf("HeatStress", X_raw, MetaData, "HeatStress")

```
5. Compile Features Present in Multiple Analyses
Combine genes and pathways found in the network, elastic net, and random forest analyses
```{r}
# === Compile genes and pathways present across multiple analyses for Biomass and HeatStress ===


# (Assuming regression and RF analyses for HeatStress have been done similarly and results saved in:)
# Gene_Lasso_Heat, Pathway_Lasso_Heat, Gene_RF_Heat, Pathways_RF_Heat

# --- 1. Network features ---
network_feats <- bind_rows(
  Gene_Network %>%
    mutate(Type = "Gene", in_network = TRUE) %>%
    dplyr::select(Feature = name, Type, network_degree = degree, in_network),
  Pathways_Network %>%
    mutate(Type = "Pathway", in_network = TRUE) %>%
    dplyr::select(Feature = name, Type, network_degree = degree, in_network)
)

# --- 2. Elastic Net features ---
enet_feats <- full_join(
  coefs_biomass %>%
    mutate(Type = ifelse(Feature %in% rownames(Plant2), "Plant", "Microbe"),
           enet_coef_biomass = Coefficient,
           in_enet_biomass = TRUE) %>%
    dplyr::select(Feature, Type, enet_coef_biomass, in_enet_biomass),
  coefs_heatstress %>%
    mutate(Type = ifelse(Feature %in% rownames(Plant2), "Plant", "Microbe"),
           enet_coef_heat = Coefficient,
           in_enet_heat = TRUE) %>%
    dplyr::select(Feature, Type, enet_coef_heat, in_enet_heat),
  by = c("Feature", "Type")
)

# --- 3. Random Forest features (top 10 per type per response) ---
rf_biomass_top <- rf_biomass %>%
  arrange(desc(Overall)) %>%
  group_by(Type) %>%
  slice_head(n = 25) %>%
  ungroup() %>%
  mutate(rf_importance_biomass = Overall, in_rf_biomass = TRUE) %>%
  dplyr::select(Feature, Type, rf_importance_biomass, in_rf_biomass)

rf_heatstress_top <- rf_heatstress %>%
  arrange(desc(Overall)) %>%
  group_by(Type) %>%
  slice_head(n = 25) %>%
  ungroup() %>%
  mutate(rf_importance_heat = Overall, in_rf_heat = TRUE) %>%
  dplyr::select(Feature, Type, rf_importance_heat, in_rf_heat)

rf_feats <- full_join(rf_biomass_top, rf_heatstress_top, by = c("Feature", "Type"))

# --- 4. Combine all features ---
compiled_feats <- full_join(network_feats, enet_feats, by = c("Feature", "Type")) %>%
  full_join(rf_feats, by = c("Feature", "Type")) %>%
  mutate(across(starts_with("in_"), ~replace_na(., FALSE)))  # Fill missing flags with FALSE

compiled_feats %>%
  summarise(
    total_features = n(),
    across(starts_with("in_"), sum)
  )

####Make a table of genes and pathways present in multiple: 
# --- 1. Create a presence matrix for all features ---
presence_matrix <- compiled_feats %>%
  dplyr::select(Feature, Type, starts_with("in_")) %>%
  mutate(across(starts_with("in_"), as.integer))  # convert TRUE/FALSE to 1/0

# --- 2. Separate Genes and Pathways ---
presence_matrix_P <- presence_matrix %>% filter(Type == "Plant")
presence_matrix_M <- presence_matrix %>% filter(Type == "Microbe")

# --- 3. Sum across all analyses per feature ---
presence_matrix_P <- presence_matrix_P %>%
  rowwise() %>%
  mutate(Total = sum(c_across(starts_with("in_")))) %>%
  ungroup()

presence_matrix_M <- presence_matrix_M %>%
  rowwise() %>%
  mutate(Total = sum(c_across(starts_with("in_")))) %>%
  ungroup()

# --- 4. Filter features present in at least two analyses ---
P_features_at_least_two <- presence_matrix_P %>% filter(Total >= 2)
M_features_at_least_two <- presence_matrix_M %>% filter(Total >= 2)

# --- 5. Print summaries ---
cat("Pathways present in >= 2 analyses (either Biomass or HeatStress):\n")
print(P_features_at_least_two)

cat("\nGenes present in >= 2 analyses (either Biomass or HeatStress):\n")
print(M_features_at_least_two)
```

A function to plot the classic GxE figure:
```{r}
plot_feature_expression <- function(Feature, X, meta_aligned) {
  # Prepare counts
  counts <- cbind(X, meta_aligned)
  counts$HeatStress <- as.numeric(counts$HeatStress)
  
  # Rename feature column to "expression"
  colnames(counts)[which(colnames(counts) == Feature)] <- "expression"
  
  # Compute means
  counts_mean <- counts %>%
    group_by(Genotype, Temp2, Soil) %>%
    dplyr::summarize(expression = mean(expression, na.rm = TRUE), .groups = "drop")
  
  # Compute standard errors
  counts_se <- counts %>%
    group_by(Genotype, Temp2, Soil) %>%
    dplyr::summarize(se = sd(expression, na.rm = TRUE) / sqrt(n()), .groups = "drop")
  
  # Merge SE into mean data
  counts_mean$se <- counts_se$se
  
  # Soil labels
  soil_labels <- c("Field" = "Field Soil", "Autoclaved" = "Autoclaved Soil")
  counts_mean$Soil <- factor(counts_mean$Soil, levels = c("Autoclaved", "Field"))
  
  # Plot
  ggplot(data = counts_mean, aes(x = Temp2, y = expression, color = Genotype)) +
    geom_point(size = 4.8) +
    geom_errorbar(aes(ymin = expression - se, ymax = expression + se), 
                  width = 0.2, linewidth = 1, alpha = .7) + 
    geom_line(aes(group = Genotype, color = Genotype), linewidth = 2, alpha = 0.9) +
    theme_classic() +
    ylab(Feature) + xlab("Temperature") +
    facet_wrap(facets = "Soil", labeller = labeller(Soil = soil_labels)) +
    scale_color_manual(values = c("goldenrod2", "forestgreen", "steelblue2")) +
    theme(text = element_text(size = 16, family = "sans"), legend.position = "none")
}

# Example usage:
# plot_feature_expression("Zm00001eb066510", X, meta_aligned)


```
```{r run function}
plot_feature_expression("Zm00001eb162740", X_raw, meta_aligned)

```