---
title: "MicrobiomeFunctionAnalysis"
author: "Nate Korth"
date: "2025-09-04"
output: html_document
---

First library packages and read in count data and metadata:
```{r setup,include = FALSE}
knitr::opts_knit$set(root.dir = "~/GERMsHeatStress/")
setwd("~/GERMsHeatStress/")
library(tidyverse)
library(ggplot2)
library(KEGGREST)
library(readxl)
library(ggpubr)
library(DESeq2)
library(vegan)
#library(Maaslin2)
library(pheatmap)
library(DESeq2)
library(data.table)
library(ALDEx2)
library(dplyr)
library(tidyr)
library(compositions)
```

This code is to analyze functional pathways (KEGG pathways) annotated from EggNOG

Step1. Read in data:
The first loop pulls both raw counts and percentages and drops into one dataframe
Next pull the raw counts and format for Maaslin anlysis

```{}
#List of all the annotation output files:
bfiles0<-list.files("./AnnotationOutputs/",pattern = "*Bacteria*",full.names = F)

df0<-NULL

for (csv in bfiles0) {
  df <- read.csv(paste0("./AnnotationOutputs/",csv), head = TRUE)
  names(df)<-c("PathwaySimple",paste0(sub("_maizeremoved_F.fastq.emapper.annotations_Bacteria.csv","",basename(csv)),"_totalCount"),paste0(sub("_maizeremoved_F.fastq.emapper.annotations_Bacteria.csv","",basename(csv)),"_PCT"))
  if (is.null(df0)) {
  df0 <- df  # First file initializes the dataframe
  } else {
    df0 <- merge(df0, df, by = "PathwaySimple", all = TRUE)  # Use all=TRUE for NA handling
  }
}

df1<-as.data.frame(t(df0))
names(df1)<-df1[1,]
df1<-df1[-1,]

df1[] <- lapply(df1, function(x) as.numeric(as.character(x)))

#Import Metadata:
MetaData<-read_excel("HotGERMs_SampleSelection.xlsx",sheet = "Sheet2")
rownames(MetaData)<-MetaData$ID2

#subset only sequenced samples
MetaData<-subset(MetaData,Sequenced=="Yes")
rownames(MetaData)<-MetaData$ID2

#Trim MetaData
MetaDataTrim<-MetaData[5:9]
MetaDataTrim$Temp<-NULL

#Pull only the percentage of reads from dataframe
df1_Percent <- df1[seq(2, nrow(df1), by = 2), ]

df1_Percent$ID2 <- gsub("_Bacteria.csv_PCT","",rownames(df1_Percent))
df1_Percent1<-merge(df1_Percent,MetaDataTrim)

df1_PercentT<-t(df1_Percent[-1])

#Pull Raw Counts for downstream analysis
pathwaycounts<-df0[, !grepl("PCT$", colnames(df0))]

# Remove unwanted pathways (a few human things that slipped through the cracks)
unwanted <- c("Melanogenesis", "Rheumatoid arthritis","Bile secretion","Meiosis  yeast","p53 signaling pathway","Photosynthesis","MAPK signaling pathway  yeast","Primary bile acid biosynthesis")
pathwaycounts <- subset(pathwaycounts, !(PathwaySimple %in% unwanted))

pathwaycountsMat <- as.matrix(pathwaycounts[-c(1)])
row.names(pathwaycountsMat)<-pathwaycounts$PathwaySimple

rownames(MetaDataTrim)<-MetaDataTrim$ID2

#Change column names:
colnames(pathwaycountsMat)<- gsub("_Bacteria.csv_totalCount","",colnames(pathwaycountsMat))

#reorder the count data to match the metadata
pathwaycountsMat <- pathwaycountsMat[, rownames(MetaDataTrim)]

#Change "NA" to 0
pathwaycountsMat[is.na(pathwaycountsMat)] <- 0
```


```{ Add for fungi}
#List of all the annotation output files:
ffiles0<-list.files("./AnnotationOutputs/",pattern = "*Fungi*",full.names = F)

df0<-NULL

for (csv in ffiles0) {
  df <- read.csv(paste0("./AnnotationOutputs/",csv), head = TRUE)
  names(df)<-c("PathwaySimple",paste0(sub("_maizeremoved_F.fastq.emapper.annotations_Fungi.csv","",basename(csv)),"_totalCount"),paste0(sub("_maizeremoved_F.fastq.emapper.annotations_Fungi.csv","",basename(csv)),"_PCT"))
  if (is.null(df0)) {
    df0 <- df  # First file initializes the dataframe
  } else {
    df0 <- merge(df0, df, by = "PathwaySimple", all = TRUE)  # Use all=TRUE for NA handling
  }
}

df1<-as.data.frame(t(df0))
names(df1)<-df1[1,]
df1<-df1[-1,]

df1[] <- lapply(df1, function(x) as.numeric(as.character(x)))

#Import Metadata:
MetaData<-read_excel("HotGERMs_SampleSelection.xlsx",sheet = "Sheet2")
rownames(MetaData)<-MetaData$ID2

#subset only sequenced samples
MetaData<-subset(MetaData,Sequenced=="Yes")
rownames(MetaData)<-MetaData$ID2

#Trim MetaData
MetaDataTrim<-MetaData[5:9]
MetaDataTrim$Temp<-NULL

#Pull only the percentage of reads from dataframe
df1_Percent <- df1[seq(2, nrow(df1), by = 2), ]

df1_Percent$ID2 <- gsub("_Fungi.csv_PCT","",rownames(df1_Percent))
df1_Percent1<-merge(df1_Percent,MetaDataTrim)

df1_PercentT<-t(df1_Percent[-1])

#Pull Raw Counts for downstream analysis
pathwaycounts<-df0[, !grepl("PCT$", colnames(df0))]

# Remove unwanted pathways (a few human things that slipped through the cracks)
unwanted <- c("Melanogenesis", "Rheumatoid arthritis","Bile secretion","p53 signaling pathway","Photosynthesis")
pathwaycounts <- subset(pathwaycounts, !(PathwaySimple %in% unwanted))

pathwaycountsMat <- as.matrix(pathwaycounts[-c(1)])
row.names(pathwaycountsMat)<-pathwaycounts$PathwaySimple

rownames(MetaDataTrim)<-MetaDataTrim$ID2

#Change column names:
colnames(pathwaycountsMat)<- gsub("_Fungi.csv_totalCount","",colnames(pathwaycountsMat))

#reorder the count data to match the metadata
pathwaycountsMat <- pathwaycountsMat[, rownames(MetaDataTrim)]

#Change "NA" to 0
pathwaycountsMat[is.na(pathwaycountsMat)] <- 0
```

Import Single gene information:
```{r}
GeneCounts<-read.csv("AnnotationOutputs/GeneCounts/Microbe_Genes_counts.csv")
GeneCounts$AllSamples<-NULL
# identify which columns are counts (everything except metadata)
count_cols <- grep("emapper.annotations", colnames(GeneCounts), value = TRUE)

sample_cols <- setdiff(colnames(GeneCounts), c("X","taxa2", "name", "pathway_EC"))

setDT(GeneCounts)
#Further combine genes that have same name/taxa
GeneCounts1 <- GeneCounts[, lapply(.SD, sum, na.rm = TRUE),
                        by = .(taxa2, name),
                        .SDcols = sample_cols]


# filter: keep rows with at least 75 counts in >= 4 samples
#GeneCounts_filt <- GeneCounts1 %>%
#  filter(rowSums(across(all_of(count_cols), ~ .x >=250)) >= 4)
GeneCounts_filt <- GeneCounts1 %>%
  filter(rowSums(across(all_of(count_cols), ~ .x >=20)) >= 2)


#If including taxa info:
GeneCounts_filt <- GeneCounts_filt %>%
  mutate(Gene = paste(taxa2, name, sep = "-")) %>%
  dplyr::select(Gene, everything())

count_mat <- GeneCounts_filt %>%
  dplyr::select(-Gene, -taxa2, -name,) %>%  # drop non-numeric metadata
  mutate(across(everything(), as.numeric)) %>% 
  as.matrix()

rownames(count_mat) <- GeneCounts_filt$Gene

#If looking at collapsed genes:
#GeneCounts_filt$x<-NULL
#count_mat <- GeneCounts_filt %>%
#  select(-name) %>%  # drop non-numeric metadata
#  mutate(across(everything(), as.numeric)) %>% 
#  as.matrix()
#rownames(count_mat) <- GeneCounts_filt$name

colnames(count_mat)<- gsub("\\.emapper\\.annotations$","",colnames(count_mat))

#Import Metadata:
MetaData<-read_excel("HotGERMs_SampleSelection.xlsx",sheet = "Sheet2")
rownames(MetaData)<-MetaData$ID2

#subset only sequenced samples
MetaData<-subset(MetaData,Sequenced=="Yes")
rownames(MetaData)<-MetaData$ID2

#Trim MetaData
MetaDataTrim<-MetaData[5:9]
MetaDataTrim$Temp<-NULL
MetaDataTrim$Biomass<-MetaData$`Biomass(Shoot)`
MetaDataTrim$HeatStress<-MetaData$HeatStress
```


```{r plot alpha diversity metrics}
# Transpose the data so that rows = samples and columns = genes/taxa
count_matrix <- t(count_mat)

# 2. Calculate Alpha Diversity Metrics
# Calculate Shannon diversity and Richness (number of observed features)
alpha_df <- data.frame(
  ID2 = rownames(count_matrix),
  Shannon = vegan::diversity(count_matrix, index = "shannon"),
  Richness = specnumber(count_matrix)
)

# 3. Merge with Metadata and Pivot Long
# Join with MetaDataTrim and pivot so we can plot both metrics easily
plot_data <- MetaDataTrim %>%
  inner_join(alpha_df, by = "ID2") %>%
  filter(Genotype != "BS") %>%
  pivot_longer(
    cols = c(Shannon, Richness),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  # Rename the levels here so the rest of the script uses the new names
  mutate(Soil = recode(Soil, "Autoclaved" = "Depleted", "Field" = "Intact"))

# 4. Generate the Plot
ggplot(plot_data, aes(x = Soil, y = Value, fill = Soil)) +
  geom_violin(trim = FALSE, alpha = 0.8, color = "black") +
  geom_boxplot(width = 0.15, outlier.shape = NA, color = "black", alpha = 0.6) +
  # Using the colors from your template
  scale_fill_manual(values = c("Depleted" = "#3C1A42FF", "Intact" = "#ED4E3EFF")) +
  theme_classic(base_size = 16) +
  # Facet by metric to show Shannon and Richness side-by-side
  facet_wrap(~ Metric, scales = "free_y") + 
  labs(
    x = "Microbiome Complexity",
    y = "Alpha Diversity",
    fill = "Soil Type",
    title = "Wilcoxon test for Microbial Functional Diversity"
  ) +
  stat_compare_means(
    comparisons = list(c("Depleted", "Intact")),
    method = "wilcox.test",
    size = 4.3,
    label = "p.format",  
    vjust = -0.71
  )

ggsave("./Revision1Files/FunctionalDiversity_AutoclavedVsField.jpeg",dpi=400)

#Make a venn diagram:
library(VennDiagram)
filtered_meta <- MetaDataTrim %>% filter(Genotype != "BS")
depleted_ids <- filtered_meta %>% filter(Soil == "Autoclaved") %>% pull(ID2)
intact_ids   <- filtered_meta %>% filter(Soil == "Field") %>% pull(ID2)

depleted_features <- rownames(count_mat)[rowSums(count_mat[, depleted_ids, drop=FALSE]) > 0]
intact_features   <- rownames(count_mat)[rowSums(count_mat[, intact_ids, drop=FALSE]) > 0]

venn_list <- list(
  Depleted = depleted_features,
  Intact = intact_features
)

venn.plot <- venn.diagram(
  x = venn_list,
  category.names = c("Depleted", "Intact"),
  filename = "./Revision1Files/Venn_Transcript_Overlap.png",
  output = TRUE,
  
  # Visual Customization matching your violin plot colors
  fill = c("#3C1A42FF", "#ED4E3EFF"),
  alpha = 0.60,
  col = "black",        # Border color
  lwd = 2,              # Border width
  
  # Text styling
  cex = 2,              # Number size
  fontface = "bold",
  cat.cex = 2,          # Category name size
  cat.default.pos = "outer",
  cat.pos = c(-20, 20), # Positioning of labels
  cat.dist = c(0.05, 0.05),
  
  main = "Overlap of Unique Transcripts",
  main.cex = 1.5
)
```
Now we'll use MaAsLin2 to pick out important pathways:

```{}
#Remove lowly expressed pathways:
pathwaycountsMatTrim<- count_mat[rowMeans(count_mat) >= 45, ]
dim(count_mat)
dim(pathwaycountsMatTrim)
# Transpose features: MaAsLin2 expects features as rows (samples) × columns (features)
features_t <- as.data.frame(t(pathwaycountsMatTrim))

# Confirm rownames match
MetaDataTrim <- as.data.frame(MetaDataTrim)
MetaDataTrim$Biomass<-as.numeric(MetaData$`Biomass(Shoot)`)
MetaDataTrim$HeatStress<-as.numeric(MetaData$HeatStress)
rownames(MetaDataTrim)<-MetaData$ID2
MetaDataTrim <- MetaDataTrim[rownames(features_t), , drop = FALSE]

#Specify interactionterms:
MetaDataTrim$Soil_Temp2 <- interaction(MetaDataTrim$Soil, MetaDataTrim$Temp2)
MetaDataTrim$Soil_Genotype <- interaction(MetaDataTrim$Soil, MetaDataTrim$Genotype)
MetaDataTrim$Temp2_Genotype <- interaction(MetaDataTrim$Temp2, MetaDataTrim$Genotype)
MetaDataTrim$Soil_Temp2_Genotype <- interaction(MetaDataTrim$Soil,MetaDataTrim$Temp2, MetaDataTrim$Genotype)

#Remove BulkSoil for maaslin analysis:
features_t$Genotype<-MetaDataTrim$Genotype
features_t_Sub<-subset(features_t,Genotype!="BS")
features_t_Sub$Genotype<-NULL
MetaDataTrim_Sub<-subset(MetaDataTrim,Genotype!="BS")

rownames(MetaDataTrim_Sub) == rownames(features_t_Sub)
rownames(MetaDataTrim) == rownames(features_t)

fit_data <- Maaslin2(
  input_data = features_t_Sub,
  input_metadata = MetaDataTrim_Sub,
  output = "maaslin2_output",
  fixed_effects = c("Soil", "Temp2", "Genotype", "Soil_Temp2","Soil_Temp2_Genotype"),
  normalization = "TSS",      # total sum scaling (adjust if already normalized)
  transform = "LOG",          # log transformation (or "NONE" if already log)
  analysis_method = "LM",     # linear model
  max_significance = 0.05,
  reference = c("Genotype,CML103"),
)

unlink("maaslin2_output/plots", recursive = TRUE)
```

Look at Maaslin Output:
```{}
# Read in the Maaslin results
sig_res <- read.delim("maaslin2_output/all_results.tsv", stringsAsFactors = FALSE)

sig_res$feature <- factor(sig_res$feature, levels = unique(sig_res$feature[order(sig_res$coef)]))

sig_resHT<-sig_res %>% filter(qval < 0.05)

# Combine metadata and value for grouping labels
sig_resHT$group <- paste(sig_resHT$metadata, sig_resHT$value, sep = ":")

# Create a unique ID for each feature-metadata combo
sig_resHT <- sig_resHT %>%
  mutate(feature_meta = paste(metadata, feature, sep = "__"))

sig_resHT <- sig_resHT %>%
  group_by(metadata) %>%
  arrange(coef, .by_group = TRUE) %>%
  mutate(feature_ordered = factor(feature_meta, levels = unique(feature_meta))) %>%
  ungroup()

#For now, exclude the interactions and genotype effect:
sig_resHT <- subset(sig_resHT, !grepl("_", metadata))
#sig_resHT <- subset(sig_resHT, !grepl("Genotype", metadata))

# Plot
ggplot(sig_resHT, aes(x = coef, y = feature_ordered, color = metadata)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = coef - stderr, xmax = coef + stderr), height = .3, size = 0.8) + 
  theme_classic() +
  labs(x = "Effect Size (Coefficient)", y = "Microbial Gene") +
  theme(text = element_text(size = 16, family = "sans"),strip.text = element_text(face = "bold"),legend.background = element_blank(),legend.position = c(0.12, 0.6),legend.box.background = element_rect(color = "black"),axis.text.y = element_blank() ) +
  scale_y_discrete(labels = function(x){ x<-gsub(".*__", "", x)
    x<-gsub("\\.", " ", x) 
    return(x)}) +  # Clean y-axis labels
  scale_color_brewer(palette = "Dark2",
                     labels = c("Soil" = "Soil Type", 
                                "Temp2" = "Temperature",
                                "Genotype" = "Genotype"))+ 
  scale_y_discrete(expand = expansion(mult = c(0, 0.01)))

ggsave("./Figures/SoilAndTempMaaslinOutput_Genes2.pdf",width=16,height=12)



```

```{r a function to plot classic gxe figure}
#Pull normalized data from Maaslin for plotting:
norm_data <- read.table("./maaslin2_output/features/filtered_data_norm.tsv", header = TRUE, row.names = 1, sep = "\t")
#norm_data<-cbind(norm_data,MetaDataTrim_Sub)

library(compositions)
#apply clr normalization
#norm_data<-subset(df1_Percent1,Genotype!="BS")
#numeric_cols <- sapply(norm_data, is.numeric)
#norm_data_clr<-as.data.frame(clr(norm_data[, numeric_cols] + 1e-6))
#norm_data <- bind_cols(norm_data[, !numeric_cols], norm_data_clr)


plot_taxa_across_datasets <- function(pathwayofinterest) {
    # Rename the selected column to "taxa" and make numeric
    df_temp <- norm_data %>% mutate(taxa = as.numeric(.data[[pathwayofinterest]]))
    
    # Mean values per group
    counts_mean <- df_temp %>%
      group_by(Genotype, Temp2, Soil) %>%
      summarise(taxa = mean(taxa, na.rm = TRUE), .groups = "drop")
    
    # SE values per group
    counts_se <- df_temp %>%
      group_by(Genotype, Temp2, Soil) %>%
      summarise(se = sd(taxa, na.rm = TRUE) / sqrt(n()), .groups = "drop")
    
    counts_mean$se <- counts_se$se
    counts_mean$Soil <- factor(counts_mean$Soil, levels = c("Autoclaved", "Field"))
    
    # Plot
    ggplot(counts_mean, aes(x = Temp2, y = taxa, color = Genotype)) +
      geom_point(size = 4.8) + 
      geom_errorbar(aes(ymin = taxa - se, ymax = taxa + se), 
                    width = 0.2, linewidth = 1, alpha = 0.7) + 
      geom_line(aes(group = Genotype), linewidth = 2, alpha = 0.9) +
      facet_wrap(~Soil, labeller = labeller(Soil = c("Field" = "Field Soil", "Autoclaved" = "Autoclaved Soil"))) +
      scale_color_manual(values = c("goldenrod2", "forestgreen", "steelblue2")) +
      labs(y = pathwayofinterest, x = "Temperature") +
      theme_classic(base_size = 14) +
      theme(legend.position = "none")
 }
```

Apply function to taxa of interest:
```{r}
plot_taxa_across_datasets("brla")
```

```{r DEseq}
MetaDataTrim2<-subset(MetaDataTrim,Genotype!="BS")
MetaDataTrim2$Group <- with(MetaDataTrim2, paste(Genotype, Soil, sep = "_"))
samples_keep <- MetaDataTrim2$ID2
count_mat <- count_mat[, colnames(count_mat) != "X", drop = FALSE]
count_mat_filtered <- count_mat[, colnames(count_mat) %in% samples_keep, drop = FALSE]

dds <- DESeqDataSetFromMatrix(countData = count_mat_filtered,
       colData = MetaDataTrim2,
       design = ~Genotype + Temp2 + Soil+Temp2:Soil)

dds <- DESeq(dds)

# Now extract log2FC for Temp2 within each Group using contrasts.
# This example constructs contrasts for each Group (GroupX at 38C vs 28C)
groups <- unique(MetaDataTrim2$Group)
res_list <- lapply(groups, function(g) {
  
  # Subset dds to only samples in this group
  dds_sub <- dds[, dds$Group == g]
  dds_sub$Temp2 <- droplevels(dds_sub$Temp2)
  design(dds_sub) <- ~ Temp2
  
  dds_sub <- DESeq(dds_sub)
  
  # Get Temp2 effect
  res <- results(dds_sub, contrast = c("Temp2", "38C", "28C"))
  
  # Return log2FC and adjusted p-values
  res_df <- as.data.frame(res)[, c("log2FoldChange", "padj")]
  colnames(res_df) <- c(paste0(g, "_LFC"), paste0(g, "_padj"))
  
  return(res_df)
})

# Combine results into a single matrix for heatmap
DE_matrix <- do.call(cbind, res_list)
lfc_mat <- DE_matrix[, grepl("_LFC$", colnames(DE_matrix)), drop = FALSE]
padj_mat <- DE_matrix[, grepl("_padj$", colnames(DE_matrix)), drop = FALSE]

# Filter: keep genes with padj < 0.05 in at least one comparison
keep_genes <- apply(padj_mat, 1, function(x) any(x < 0.01, na.rm = TRUE))
lfc_filtered <- lfc_mat[keep_genes, ]

# Convert to matrix
lfc_matrix <- as.matrix(lfc_filtered)


# Then pheatmap(res_mat)
pheatmap(lfc_matrix,
         scale = "row",                 # row-scale expression
         clustering_method = "ward.D2",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         show_colnames = TRUE,
         show_rownames = FALSE,         # hide row names if too many
         fontsize_col = 10,
         color = colorRampPalette(c("darkorchid4", "grey99", "indianred2"))(50))
```

Check PCA
```{r}
mat <- as.matrix(GeneCounts_filt[, sample_cols, with = FALSE])
rownames(mat) <- paste(GeneCounts_filt$taxa, GeneCounts_filt$name, sep = "_")
colnames(mat)<- gsub("\\.emapper\\.annotations$","",colnames(mat))
# --- CLR transform ---
# Transpose first so rows = samples, cols = features
mat_t <- t(mat)
# Convert counts to relative abundance per sample
mat_rel <- mat_t / rowSums(mat_t)

mat_clr <- clr(mat_rel)   # compositions::clr expects samples x features

# --- PCA ---
pca <- prcomp(mat_clr, center = TRUE, scale. = FALSE)

# --- Extract scores for plotting ---
pca_df <- as.data.frame(pca$x)

# Bring in sample metadata (e.g., Genotype, Soil, Temp2)
pca_df$ID2 <- rownames(pca_df)
pca_df <- merge(pca_df, MetaDataTrim2, by = "ID2")

pca_df$Biomass<-as.numeric(pca_df$Biomass)

# --- Plot ---
ggplot(pca_df, aes(x = PC1, y = PC2, color = Soil, shape = Temp2)) +
  geom_point(size = 4.5,alpha=.85) +
  theme_classic(base_size = 19) +
  scale_color_manual(values = c("#400F73FF", "#FA815FFF", "steelblue2")) +
  #scale_color_manual(values = c("darkgreen", "green", "yellow","orange","red")) +
  #scale_color_viridis_c(option = "plasma") +
  labs(x = paste0("PC1 (", round(100*summary(pca)$importance[2,1], 1), "%)"),
       y = paste0("PC2 (", round(100*summary(pca)$importance[2,2], 1), "%)"),
       shape = "Temperature")
ggsave("../Figures/MicrobeGenesPCA.pdf",width=9,height=4.4)
```

Using ALDEx2 let's try a different DE analysis:
```{}
# Make sure metadata matches count matrix
MetaDataTrim2 <- subset(MetaDataTrim, Genotype != "BS")
MetaDataTrim2$Group <- paste(MetaDataTrim2$Genotype, MetaDataTrim2$Soil, sep = "_")
samples_keep <- intersect(colnames(count_mat), MetaDataTrim2$ID2)

count_sub <- count_mat[, samples_keep, drop = FALSE]
meta_sub  <- MetaDataTrim2 %>% filter(ID2 %in% samples_keep) %>%
  arrange(match(ID2, colnames(count_sub)))
rownames(meta_sub) <- meta_sub$ID2

stopifnot(all(colnames(count_sub) == rownames(meta_sub)))

mm <- model.matrix(~ Temp2 + Soil + Genotype+Temp2:Genotype, MetaDataTrim2)

# CLR transform
x.glm <- aldex.clr(count_sub,mm, mc.samples = 128, denom = "all", verbose = TRUE)
glm.test <- aldex.glm(x.glm, mm, fdr.method='holm')
glm.eff<- aldex.glm.effect(x.glm)

```
```{r Make heat map}
samples_keep <- intersect(colnames(count_mat), MetaDataTrim2$ID2)
count_sub <- count_mat[, samples_keep, drop = FALSE]
meta_sub  <- MetaDataTrim2 %>% filter(ID2 %in% samples_keep) %>%
  arrange(match(ID2, colnames(count_sub)))
rownames(meta_sub) <- meta_sub$ID2
meta_sub$Group <- with(meta_sub, paste(Genotype, Soil, sep = "_"))
groups <- unique(meta_sub$Group)
group_results <- list()


#g <- groups[1] 

for (g in groups) {
  # Subset samples in this group
  samp_ids <- meta_sub$ID2[meta_sub$Group == g]

  # Keep only samples present in count_sub
  samp_ids <- intersect(samp_ids, colnames(count_sub))
  if (length(samp_ids) == 0) next  # skip if no samples

  count_g <- count_sub[, samp_ids, drop = FALSE]
  meta_g  <- meta_sub[samp_ids, , drop = FALSE]

  meta_g <- meta_sub[meta_sub$ID2 %in% samp_ids, , drop = FALSE]
  
  mm_g <- model.matrix(~ Temp2, data = meta_g)
  
  # Run ALDEx2 
  x_g <- aldex.clr(count_g, mm_g, mc.samples = 136, denom = "all", verbose = FALSE)
  
  glm.test_g <- aldex.glm(x_g, mm_g, fdr.method='BH') 
  
  glm.eff_g <- aldex.glm.effect(x_g) 
  # Extract Temp2 effect 
  eff_df <- glm.eff_g[["Temp238C"]] %>%
    as.data.frame() %>%
    rownames_to_column("feature")
  
stats_df <- glm.test_g %>%
  as.data.frame() %>%
  tibble::rownames_to_column("feature") %>%
  dplyr::select(
    feature,
    Temp2.Est  = !!sym("Temp238C:Est"),
    Temp2.pval = !!sym("Temp238C:pval"),
    Temp2.padj = !!sym("Temp238C:pval.padj")
  )
  
merged <- left_join(eff_df, stats_df, by = "feature") %>%
  dplyr::select(feature, diff.btw, Temp2.Est,Temp2.pval, Temp2.padj) %>%
  dplyr::rename(
    !!paste0(g, "_Est")  := Temp2.Est,
    !!paste0(g, "_pval") := Temp2.pval,
    !!paste0(g, "_padj") := Temp2.padj
  )
  
  group_results[[g]] <- merged
}

merged_all <- Reduce(function(x, y) full_join(x, y, by = "feature"), group_results)

# Grab the effect size (Est) columns
lfc_mat <- merged_all %>%
  dplyr::select(feature, ends_with("_Est"))

# Grab padj columns
padj_mat <- merged_all %>%
  dplyr::select(feature, ends_with("_padj"))

# Keep genes with padj < 0.05 in at least one group
keep_genes <- apply(padj_mat[,-1], 1, function(x) any(x < 0.05, na.rm = TRUE))

lfc_filtered <- lfc_mat[keep_genes, ]

# Convert to matrix
lfc_matrix1 <- as.matrix(lfc_filtered[,-1])
rownames(lfc_matrix1) <- lfc_filtered$feature

#Save Results:
write.csv(lfc_matrix1,"ALDEx2Out_Temp.csv")

desired_order <- c("CML52_Field_Est", "CML52_Autoclaved_Est",
                   "CML103_Field_Est", "CML103_Autoclaved_Est",
                   "SAP_Field_Est", "SAP_Autoclaved_Est")
lfc_matrix1 <- lfc_matrix1[, intersect(desired_order, colnames(lfc_matrix1)), drop = FALSE]


# -------------------
# 5. Heatmap
# -------------------

pheatmap(
  lfc_matrix1,
  scale = "row",
  clustering_method = "ward.D2",
  clustering_distance_rows = "euclidean",
  cluster_cols = FALSE,  
  show_colnames = TRUE,
  show_rownames = FALSE,
  treeheight_col = 0,
  fontsize_col = 10,
  color = colorRampPalette(c("#3C1A42FF", "grey99", "#ED4E3EFF"))(50),
  filename="../Figures/ALDEX2HeatMap_MicrobeGenes.tiff",width = 6,height = 10
)


```

Import Collapsed Pathway information:
```{r}
GeneCounts<-read.csv("../AnnotationOutputs/GeneCounts/Microbe_Genes_collapsed_counts.csv")
GeneCounts$AllSamples<-NULL
# identify which columns are counts (everything except metadata)
count_cols <- grep("emapper.annotations", colnames(GeneCounts), value = TRUE)

sample_cols <- setdiff(colnames(GeneCounts), c("X","taxa2", "name", "desc"))

setDT(GeneCounts)
#Further combine genes that have same name/taxa
#GeneCounts1 <- GeneCounts[, lapply(.SD, sum, na.rm = TRUE),
#                        by = .(taxa2, name),
#                        .SDcols = sample_cols]


# filter: keep rows with at least 75 counts in >= 4 samples
GeneCounts_filt <- GeneCounts1 %>%
  filter(rowSums(across(all_of(count_cols), ~ .x >=250)) >= 4)

#If including taxa info:
GeneCounts_filt <- GeneCounts_filt %>%
  mutate(Gene = paste(taxa2, name, sep = "-")) %>%
  dplyr::select(Gene, everything())

count_mat <- GeneCounts_filt %>%
  dplyr::select(-Gene, -taxa2, -name,) %>%  # drop non-numeric metadata
  mutate(across(everything(), as.numeric)) %>% 
  as.matrix()

rownames(count_mat) <- GeneCounts_filt$Gene

#If looking at collapsed genes:
#GeneCounts_filt$x<-NULL
#count_mat <- GeneCounts_filt %>%
#  select(-name) %>%  # drop non-numeric metadata
#  mutate(across(everything(), as.numeric)) %>% 
#  as.matrix()
#rownames(count_mat) <- GeneCounts_filt$name

colnames(count_mat)<- gsub("\\.emapper\\.annotations$","",colnames(count_mat))
```


Same type of thing, just using pathways:

```{r}
PathCounts<-read.csv("../AnnotationOutputs/GeneCounts/Microbe_Pathways_collapsed_EC_counts.csv")
PathCounts$X<-NULL
# identify which columns are counts (everything except metadata)
count_cols <- grep("emapper.annotations", colnames(PathCounts), value = TRUE)

sample_cols <- setdiff(colnames(PathCounts), c("pathway", "desc"))

setDT(PathCounts)

PathCounts_filt <- PathCounts %>%
  filter(!grepl("cancer|disease|bile acid|Photosynthesis|Drug metabolism|Human", desc, ignore.case = TRUE))

# filter: keep rows with at least 75 counts in >= 4 samples
PathCounts_filt <- PathCounts_filt %>%
  filter(rowSums(across(all_of(count_cols), ~ .x >=250)) >= 4)

count_mat <- PathCounts_filt %>%
  dplyr::select(-pathway, -desc,) %>%  # drop non-numeric metadata
  mutate(across(everything(), as.numeric)) %>% 
  as.matrix()

rownames(count_mat) <- PathCounts_filt$desc

#If looking at collapsed genes:
#GeneCounts_filt$x<-NULL
#count_mat <- GeneCounts_filt %>%
#  select(-name) %>%  # drop non-numeric metadata
#  mutate(across(everything(), as.numeric)) %>% 
#  as.matrix()
#rownames(count_mat) <- GeneCounts_filt$name

colnames(count_mat)<- gsub("\\.emapper\\.annotations$","",colnames(count_mat))

```

Run AlDEx2
```{r}
# Keep only desired samples
samples_keep <- MetaDataTrim$ID2[MetaDataTrim$Genotype != "BS"]
count_sub <- count_mat[, samples_keep, drop = FALSE]
meta_sub  <- MetaDataTrim %>%
  filter(ID2 %in% samples_keep) %>%
  arrange(match(ID2, colnames(count_sub))) %>%
  as.data.frame()  # ensure data.frame
rownames(meta_sub) <- meta_sub$ID2

stopifnot(all(colnames(count_sub) == rownames(meta_sub)))

# Make sure factors
meta_sub$Genotype <- factor(meta_sub$Genotype)
meta_sub$Soil     <- factor(meta_sub$Soil)
meta_sub$Temp2    <- factor(meta_sub$Temp2)
meta_sub$Genotype <- relevel(meta_sub$Genotype, ref = "SAP")   # CML103 as reference

# Run ALDEx2 GLM with formula (no model.matrix!)
full_mm <- model.matrix(~ Genotype + Soil + Temp2 + Temp2:Soil,
                        data = meta_sub)

x.glm <- aldex.clr(count_sub, full_mm, mc.samples = 228, denom = "all", verbose = TRUE)
glm.test <- aldex.glm(x.glm, full_mm, fdr.method='BH')
glm.eff  <- aldex.glm.effect(x.glm)

# Calculate CML52 vs CML103 difference manually
glm.test$`GenotypeCML52_vs_CML103:Est`  <- glm.test$`GenotypeCML52:Est` - glm.test$`GenotypeCML103:Est`
glm.test$`GenotypeCML52_vs_CML103:SE`   <- sqrt(glm.test$`GenotypeCML52:SE`^2 + glm.test$`GenotypeCML103:SE`^2)
glm.test$`GenotypeCML52_vs_CML103:pval` <- 2 * pnorm(-abs(glm.test$`GenotypeCML52_vs_CML103:Est` / glm.test$`GenotypeCML52_vs_CML103:SE`))




# Subset columns for three-way interactions
main_effect_cols <- c(
  "GenotypeCML52:Est", "GenotypeCML103:Est","GenotypeCML52_vs_CML103:Est",
  "SoilField:Est",
  "Temp238C:Est",
  "SoilField:Temp238C:Est")

se_cols <- sub(":Est$", ":SE", main_effect_cols)
p_cols  <- sub(":Est$", ":pval", main_effect_cols)




# Top 10 pathways by min p-value across selected effects
#glm.test$min_pval <- apply(glm.test[, p_cols], 1, min, na.rm = TRUE)
#top10_idx <- order(glm.test$min_pval)[1:50]
#top10 <- glm.test[top10_idx, ]

glm.test$sum_abs_effect <- rowSums(abs(glm.test[, main_effect_cols]), na.rm = TRUE)

# Pick top 15 features by total effect magnitude
top_idx <- order(glm.test$sum_abs_effect, decreasing = TRUE)[1:16]
top15 <- glm.test[top_idx, ]

# Convert to long format
plot_df <- top15 %>%
  dplyr::select(all_of(c(main_effect_cols, se_cols))) %>%
  mutate(Pathway = rownames(.)) %>%
  pivot_longer(
    cols = -Pathway,
    names_to = c("Effect", ".value"),
    names_pattern = "(.*):(Est|SE)"
  )

# Clean pathway names
plot_df$Pathway <- sub(";.*$", "", plot_df$Pathway)

# Set custom order for Effect
plot_df$Effect <- factor(plot_df$Effect, levels = c(
  "SoilField",
  "Temp238C",
  "GenotypeCML52",
  "GenotypeCML103",
  "GenotypeCML52_vs_CML103",
  "SoilField:Temp238C"
))


# Plot
ggplot(plot_df, aes(x = Est, y = reorder(Pathway, Est), fill = Effect)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(xmin = Est - SE, xmax = Est + SE),
                position = position_dodge(width = 0.8), width = 0.2) +
  labs(x = "Effect Size", y = "Pathway", fill = "Effect") +
  theme_classic() +scale_fill_manual(values = c("#000004FF","#C63C74FF","#FECD90FF","#F9785DFF","#7C2382FF","#331067FF"))+
  theme(axis.text.y = element_text(size = 13))
ggsave("../Figures/ALDEX2Bar_MicrobePaths.pdf",width = 12,height=8)

```

